\documentclass[11pt]{article}

% EMNLP 2026 Demo Track is single-blind: authors are named.
% [preprint] = non-anonymous + page numbers (no line numbers).
% Switch to [final] for the camera-ready.
\usepackage[preprint]{acl}

% Standard package includes
\usepackage{times}
\usepackage{latexsym}

% For proper rendering and hyphenation of words containing Latin characters.
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{inconsolata}

% Figures and diagrams.
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{positioning, arrows.meta, calc}

% Tables and math.
\usepackage{booktabs}
\usepackage{colortbl}
\usepackage{amsmath}
\usepackage{amssymb}

% Inline author note for proposed design choices that need confirmation.
% xcolor is already loaded by acl.sty.
\newcommand{\designnote}[1]{\textcolor{red}{\textit{[TODO: #1]}}}
\definecolor{featureyes}{RGB}{220,232,216}
\definecolor{featurepartial}{RGB}{242,226,184}
\definecolor{featureno}{RGB}{237,211,207}
\DeclareRobustCommand{\halfcheck}{\ensuremath{\checkmark\mkern-10mu/}}
\DeclareRobustCommand{\featurelegendbox}[2]{\begingroup\setlength{\fboxsep}{1pt}\colorbox{#1}{\makebox[1.15em]{#2}}\endgroup}
\newcommand{\yesfeature}{\cellcolor{featureyes}\ensuremath{\checkmark}}
\newcommand{\partialfeature}{\cellcolor{featurepartial}\halfcheck}
\newcommand{\nofeature}{\cellcolor{featureno}\ensuremath{\times}}

\title{Humanly: Process Evidence for Human-AI Writing}

% EMNLP 2026 Demo Track is single-blind: fill in real names and affiliations.
\author{First Author \\
  Affiliation / Address line 1 \\
  \texttt{email@domain} \\\And
  Second Author \\
  Affiliation / Address line 1 \\
  \texttt{email@domain} \\}

\begin{document}
\maketitle

\begin{abstract}
Large language models make it increasingly difficult to infer how a piece of writing was produced from the final text alone. This is a policy problem, not only a detection problem: in classrooms, peer review, and personal certification, AI assistance may be allowed, limited, or prohibited depending on the task. We present \textsc{Humanly}, a provenance-first writing platform that records how a draft comes together and turns that record into verifiable process evidence. Writers draft in a tracked editor; task owners configure AI access, paste rules, character limits, and time limits; and the system records typing, paste, focus, selection, and AI-assist events. At completion, \textsc{Humanly} issues a certificate and verification link that connect the final text to typed/pasted statistics, AI-use logs, and replayable writing history. We compare \textsc{Humanly} with final-text AI detectors and process-replay systems on the policy-compliance questions they can and cannot answer, then outline an evaluation through detector stress tests and a human study of writer usability and reader interpretation of certificate evidence. The system is deployed at \url{https://app.writehumanly.net/}.
\end{abstract}

\section{Introduction}
\label{sec:intro}

Large language models have changed the status of writing evidence. A fluent final draft no longer tells an instructor whether a student composed the argument, a program chair whether a reviewer engaged with a paper, or a reader whether a writing sample reflects a person's own process. The question is often stated as ``did a human or AI write this?'' Yet in practice the sharper question is policy-compliance: what assistance was allowed for this task, and what evidence shows how the text was produced?

Most existing responses operate after the fact. Text detectors judge whether the final document resembles model output \citep{mitchell2023detectgpt, hans2024binoculars, verma2024ghostbuster}. This can be useful for some fully generated text, but it cannot distinguish many mixed workflows: a human draft polished by AI, a non-native English writer falsely flagged as AI-generated \citep{liang2023gpt}, or an AI draft paraphrased by a humanizer. Process-replay tools improve the situation by showing document history, but replay alone does not necessarily define a task policy, log sanctioned AI use, or issue portable evidence that a writer can share with a reader.

\textsc{Humanly} starts from a different principle: \emph{process beats prediction}. Instead of inferring authorship from style, Humanly records the writing process as it happens. A task owner can configure AI access, paste rules, time limits, and character limits before writing begins. A writer then drafts inside a tracked editor where typed text, paste events, focus changes, selections, and AI interactions are recorded. At the end, Humanly generates a certificate and verification link that connect the final draft to a process record.

This framing allows AI to be treated as a governed part of writing rather than as a binary contaminant. Humanly can be used when AI is disabled, when only AI polishing is allowed, when only chat assistance is allowed, or when full AI assistance is allowed. In each case, the relevant evidence is not whether the final prose ``sounds AI-generated,'' but whether the process matches the rules set for the task.

Our contributions are:
\begin{enumerate}
  \item a deployed, AI-native writing platform that captures write-time provenance and issues verifiable certificates;
  \item configurable writing environments for personal documents and assigned tasks, including four AI-access modes, paste policy, character limits, and time limits;
  \item a feature comparison that contrasts Humanly with final-text AI detectors and document-replay tools on policy-compliance questions;
  \item an evaluation plan covering detector stress tests, writer usability, policy clarity, and reader interpretation of certificate evidence.
\end{enumerate}
\designnote{Replace the third and fourth contributions with actual evaluation findings once data collection is complete.}

\section{System Architecture}
\label{sec:architecture}

\textsc{Humanly} is a deployed web platform with three production surfaces: a user portal for writers, an admin dashboard for task owners, and an API/tracker host. The implementation is a TypeScript monorepo: an Express backend with PostgreSQL/TimescaleDB and Redis, a Next.js admin dashboard, a Next.js user portal, a Lexical editor package, a standalone tracker package, and shared validators/types. Figure~\ref{fig:workflow} sketches the current data path.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  every node/.style={font=\small},
  box/.style={draw, rectangle, rounded corners=2pt, align=center, inner sep=4pt, minimum height=7mm},
  core/.style={box, fill=gray!12},
  arr/.style={-Stealth, thick}
]
  \node[box] (admin) {Admin Dashboard\\\scriptsize task owner};
  \node[box, right=8mm of admin] (user) {User Portal\\\scriptsize writer};

  \node[core, below=10mm of admin, xshift=14mm, minimum width=46mm] (editor)
    {Lexical Editor + Tracker\\\scriptsize typing\,$\cdot$\,paste\,$\cdot$\,focus\,$\cdot$\,select\,$\cdot$\,AI};
  \node[core, below=2mm of editor, minimum width=46mm] (ai)
    {AI Policy\\\scriptsize off\,$\cdot$\,polish\,$\cdot$\,chat\,$\cdot$\,full};

  \node[core, below=8mm of ai, minimum width=58mm] (backend)
    {Backend\,$\cdot$\,TimescaleDB\,$\cdot$\,Redis\,$\cdot$\,WebSocket};

  \node[box, below=10mm of backend.south, xshift=-15mm] (cert) {Certificate\\\scriptsize PDF + JSON + verify};
  \node[box, below=10mm of backend.south, xshift=15mm] (review) {Review Surfaces\\\scriptsize logs + replay};

  \draw[arr] (admin.south) -- node[midway, left=1pt] {\scriptsize task policy} (admin.south |- editor.north);
  \draw[arr] (user.south) -- (user.south |- editor.north);
  \draw[arr] (ai.south) -- node[midway, right=1pt] {\scriptsize events} (backend.north);
  \draw[arr] (backend.south -| cert.north) -- (cert.north);
  \draw[arr] (backend.south -| review.north) -- (review.north);
\end{tikzpicture}
\caption{Humanly records the writing process at the editor surface, stores events and document state in the backend, and produces certificate, verification, log, and replay surfaces for later review.}
\label{fig:workflow}
\end{figure}

\paragraph{Writing environments.}
Each document is governed by an environment configuration. The configuration records whether the task is personal or admin-assigned, whether instruction PDFs are present, which AI provider and model are allowed, how much AI use is permitted, whether paste is allowed or blocked, whether start/end windows and a writing timer apply, and whether minimum or maximum character counts are enforced. The current AI-access modes are \textsc{off}, \textsc{polish}, \textsc{chat}, and \textsc{full}. In \textsc{polish}, only selection-level actions such as fixing grammar, improving writing, simplifying, or formalizing are available. In \textsc{chat}, only the agent chat is available. In \textsc{full}, both are available. In \textsc{off}, AI is disabled.

\paragraph{Editor and recording layer.}
Writers draft in a Lexical-based editor instrumented with provenance capture. The system records typing, paste, focus/blur, selection, and AI-assist events as part of the writing session. When AI is enabled, Humanly records both agent-chat interactions and selection-level AI actions, including whether the writer accepts or rejects generated text. This makes AI collaboration part of the process record rather than an invisible side channel.

\paragraph{Task and enrollment layer.}
Admin-assigned tasks are created in the admin dashboard and can be joined through authenticated invite-code enrollment or public share links. A task owner can decide whether a public link allows guest submissions. Public writers must start a Humanly document before submitting; direct public submission is disabled. Signed-in public writers enter the normal task/certificate flow, while guest sessions are mapped to synthetic guest users so the same editor and certificate pipeline can run.

\paragraph{Storage and review layer.}
The backend stores document state, event streams, task enrollments, submissions, AI chat sessions, and AI selection actions. PostgreSQL/TimescaleDB supports high-frequency event storage, while WebSocket channels support live or near-live review surfaces in the admin workflow. \designnote{Verify which live-monitoring surfaces should be shown in the final screenshot set.}

\paragraph{Certificate and verification layer.}
At completion, Humanly generates a certificate containing the document title, generated time, certificate identifier, verification token, typed and pasted character counts, event counts, editing time, and AI-use statistics. Certificates can be exported as JSON and PDF and can be verified through a public URL. When enabled, certificates can include the full text, edit history, replay/log surfaces, and an access-code gate.

\section{Workflows and Use Cases}
\label{sec:workflows}

Humanly supports three primary workflows. Each is organized around the same principle: the relevant artifact is not only the final text, but also the process evidence attached to it.

\subsection{Peer Review of Academic Papers}

Peer review is a natural policy-compliance setting. Venues increasingly need to specify whether reviewers may use LLMs for grammar, polishing, summarization, or substantive review. Final-text detectors cannot tell whether a review reflects a reader's own engagement with the paper or whether an external model produced the main judgment. In Humanly, a program chair can create an assigned review task, attach instruction files or paper context, configure the permitted AI mode, and ask reviewers to write inside the tracked workspace. The resulting certificate gives the chair a process trail: writing duration, typed and pasted content, AI interactions, and replayable history.

\subsection{Classroom Writing Assignments}

Instructors face the same policy problem at a larger scale. Some courses ban AI assistance; others allow grammar correction, brainstorming, or full collaboration. Humanly lets instructors configure those rules before writing starts. Students then write in an environment where the interface itself reflects the permitted mode. On submission, the instructor can review the final text together with typed/pasted statistics, paste-policy evidence, AI-use logs, and certificate metadata. This shifts the interaction from accusation and defense to a shared process record.

\subsection{Personal Writing Certification}

Writers can also use Humanly without an admin. A writer creates a personal document, drafts with or without AI, and generates a certificate when they want to show how the document was produced. This supports writing samples, editorial workflows, and situations where a writer wants a portable receipt of process rather than a detector score.

\designnote{Replace this paragraph with three final screenshots: peer-review/admin task setup, user writing surface, and certificate/verification view.}

\section{Comparison with Existing Systems}
\label{sec:comparison}

Humanly sits between two existing responses to AI-era writing authenticity: final-text detectors and process/replay systems. The comparison in this section is deliberately separate from the evaluation in Section~\ref{sec:evaluation}. Here we ask what each class of system is designed to answer, and where Humanly adds a different workflow.

\subsection{Final-Text Detectors}

Final-text detectors such as GPTZero,\footnote{\url{https://gptzero.me/}} Pangram,\footnote{\url{https://www.pangram.com/}} Originality.AI,\footnote{\url{https://originality.ai/}} and Copyleaks\footnote{\url{https://copyleaks.com/ai-content-detector}} evaluate the submitted text after writing is complete. This can be useful when a task bans AI-generated text outright and the output is substantially machine-generated. However, final-text detection does not directly observe the writing process. It therefore cannot distinguish policy-compliant AI use from policy-violating AI use when a task permits some forms of assistance.

This limitation is structural rather than only a matter of detector accuracy. A human-written draft polished by AI, a human-written draft translated by AI, and an AI-generated answer paraphrased by a humanizer may all produce fluent final prose. Conversely, human prose can be formal, template-like, or non-native in ways that resemble model output \citep{liang2023gpt}. In these cases, the policy question is not simply whether the text looks AI-generated. The relevant question is whether the process matched the assistance allowed for the task.

\subsection{Process and Replay Systems}

Process and replay tools move closer to the Humanly framing because they expose evidence about how a document changed over time. Turnitin Clarity provides writing reports and AI chat activity in an assignment workflow.\footnote{\url{https://guides.turnitin.com/hc/en-us/articles/36916426919949-Reviewing-the-Writing-Report}} Grammarly Authorship categorizes typed, pasted, AI-generated, AI-rephrased, and Grammarly-edited text across supported writing surfaces.\footnote{\url{https://support.grammarly.com/hc/en-us/articles/29548735595405-About-Authorship}} GPTZero Writing Reports combine detector-adjacent authorship reports with Google Docs writing-process evidence.\footnote{\url{https://support.gptzero.me/articles/7001890416-what-is-the-google-docs-writing-report-for-origin}} Google Docs replay tools such as Draftback,\footnote{\url{https://draftback.com/}} Brisk Inspect Writing,\footnote{\url{https://help.briskteaching.com/hc/en-us/articles/39310227756436-How-To-Inspect-Writing}} Integrito,\footnote{\url{https://integrito.ai/}} WritingTrace,\footnote{\url{https://writingtrace.com/}} and PaperTrail Inspect\footnote{\url{https://papertrailacademic.com/inspect/}} reconstruct document history, paste events, sessions, or process reports from host-editor data.

These systems show that replay itself is not the unique contribution. Humanly instead integrates four pieces that are usually separated: a configurable writing environment, native AI-use logging, assigned-task enrollment, and certificate analytics beyond replay. Table~\ref{tab:feature-comparison} summarizes this feature comparison. We mark a feature as partial when public documentation shows related support in a narrower product context. Examples include Google Docs extensions and assignment-specific reports.

\begin{table*}[t]
\scriptsize
\centering
\setlength{\tabcolsep}{2.7pt}
\begin{tabular}{p{0.19\textwidth}p{0.24\textwidth}cccccccc}
\toprule
\multicolumn{2}{c}{} & \multicolumn{2}{c}{Environment} & \multicolumn{2}{c}{Governance} & \multicolumn{2}{c}{Capture} & \multicolumn{2}{c}{Evidence} \\
\cmidrule(lr){3-4}\cmidrule(lr){5-6}\cmidrule(lr){7-8}\cmidrule(lr){9-10}
System & Type & Native & Flex Env & AI Policy & Assigned & Event & AI Log & Replay & Cert. \\
\midrule
\textsc{Humanly} & Native provenance platform & \yesfeature & \yesfeature & \yesfeature & \yesfeature & \yesfeature & \yesfeature & \yesfeature & \yesfeature \\
Turnitin Clarity & Assignment writing workspace & \yesfeature & \partialfeature & \partialfeature & \yesfeature & \partialfeature & \yesfeature & \yesfeature & \partialfeature \\
Grammarly Authorship & Cross-surface attribution & \partialfeature & \partialfeature & \partialfeature & \nofeature & \partialfeature & \partialfeature & \yesfeature & \yesfeature \\
GPTZero Writing Reports & Detector + Google Docs report & \nofeature & \nofeature & \nofeature & \nofeature & \partialfeature & \nofeature & \yesfeature & \partialfeature \\
Draftback & Google Docs revision replay & \nofeature & \nofeature & \nofeature & \nofeature & \partialfeature & \nofeature & \yesfeature & \partialfeature \\
Brisk Inspect Writing & Google Docs classroom replay & \nofeature & \nofeature & \nofeature & \partialfeature & \partialfeature & \nofeature & \yesfeature & \nofeature \\
Integrito & Google Docs activity report & \nofeature & \nofeature & \nofeature & \partialfeature & \partialfeature & \nofeature & \partialfeature & \partialfeature \\
WritingTrace & Google Docs forensic replay & \nofeature & \nofeature & \nofeature & \nofeature & \partialfeature & \nofeature & \yesfeature & \yesfeature \\
PaperTrail Inspect & Google Docs process report & \nofeature & \nofeature & \nofeature & \nofeature & \partialfeature & \nofeature & \yesfeature & \yesfeature \\
\bottomrule
\end{tabular}
\caption{Feature comparison for process and replay systems. Legend: \featurelegendbox{featureyes}{\(\checkmark\)} supported; \featurelegendbox{featurepartial}{\(\halfcheck\)} partial or narrower; \featurelegendbox{featureno}{\(\times\)} not found. Native = controlled writing workspace; Flex Env = configurable writing rules; Event = fine-grained process events beyond coarse playback; Cert. = certificate or report analytics beyond replay.}
\label{tab:feature-comparison}
\end{table*}

The table reflects a narrow claim. Turnitin Clarity is the strongest education comparator because it combines an assignment writing workspace, writing playback, and AI chat activity. Grammarly Authorship is a strong attribution comparator because it labels typed, pasted, AI-generated, and AI-rephrased text. WritingTrace and PaperTrail Inspect are close philosophical comparators because they emphasize process evidence rather than detector probability. GPTZero is notable because it advertises AI reviewing, writing replay, plagiarism checks, and AI detection as a classroom toolkit.\footnote{\url{https://gptzero.me/news/turnitin-vs-brisk/}} However, the public-facing documentation does not establish a single write-time provenance record that links assistant actions, detector results, replay, task policy, and verification. The Humanly contribution is not that these capabilities are absent elsewhere. Rather, Humanly produces policy setup, AI access modes, write-time event capture, participant enrollment, and certificate analytics from the same provenance pipeline.

\subsection{Synthesis}

Final-text detectors answer what the final text resembles. Process and replay tools answer part of what happened inside a document. Humanly is designed around whether the writing process complied with the policy set for the task. This distinction matters because AI assistance is no longer a binary contaminant. The same course, venue, or organization may allow grammar correction, prohibit generation, permit chat-based explanation, or allow full collaboration. A useful writing-authenticity system must therefore record both the policy and the process.

\section{Evaluation}
\label{sec:evaluation}

We evaluate Humanly from two complementary angles. First, a final-text detector stress test measures where text-only systems fail as policy-compliance evidence. Second, a human study evaluates whether writers and readers can use process evidence from Humanly in realistic workflows. Both evaluations treat the Humanly certificate as evidence for downstream judgment, not as an automatic verdict.

\subsection{Final-Text Detector Stress Test}
\label{sec:detector-eval}

The detector benchmark uses situation cases rather than only aggregate accuracy. We define one policy for the first benchmark: human writing, AI polish, and AI translation are allowed, while substantive AI generation of final content is not allowed. Under this policy, human-origin or policy-compliant cases are the negative class for detector-style binary scoring, and AI-origin or policy-violating cases are the positive class.

\begin{table}[t]
\small
\centering
\setlength{\tabcolsep}{4pt}
\begin{tabular}{p{0.28\columnwidth} p{0.58\columnwidth}}
\toprule
Risk class & Situation cases \\
\midrule
False positive risk &
Human original; human + AI polish; human + AI translation; human-written AI-style prose. \\
False negative risk &
Direct AI-generated; AI-obfuscated through humanizer, paraphrase, or ``sound human'' prompting; AI cross-lingual transform; AI-generated + light human edits. \\
\bottomrule
\end{tabular}
\caption{Planned detector stress-test matrix. Cases are labelled by policy compliance, not by whether Humanly predicts authorship. The final paper will replace or extend this table with detector outputs and confusion matrices.}
\label{tab:detector-cases}
\end{table}

For each detector, we will record raw scores, native labels, thresholded binary labels, and failure modes by situation. The report will emphasize false positives on policy-compliant text and false negatives on policy-violating text, because both make policy-compliance judgment unreliable. \designnote{Run GPTZero, Pangram, Originality.AI, and Copyleaks if accessible. Replace this plan with results once available.}

\subsection{Human Study}
\label{sec:study}

The human study evaluates Humanly itself. We plan to measure three questions. First, can writers use the tracked workspace naturally? Second, is the configured policy clear during writing? Third, can readers use certificate evidence to make better policy-compliance judgments? We will use tasks drawn from peer review, classroom writing, online survey responses, and social-media-style posts. Peer review is the flagship setting because it combines writing quality, document understanding, confidentiality, and explicit AI-use policy.

\paragraph{Writer-Side Task.}
Participants write short documents in Humanly under selected environment policies: \textsc{off}, \textsc{polish}, \textsc{chat}, and \textsc{full}. We will collect usability ratings, perceived friction, policy clarity, and behavioral statistics from the recorded trajectories. Survey questions ask whether the interface is easy to use and whether participants would use Humanly in real settings. They also ask how participants feel about writing under process recording.

\paragraph{Reader-Side Task.}
Readers inspect writing evidence under different information conditions: final text only, final text plus detector score, and final text plus Humanly certificate/replay. We will measure accuracy against known workflow labels, confidence, time-to-judgment, perceived fairness, and which evidence readers report using.

\paragraph{Hackability and Trace Taxonomy.}
We will separately evaluate adversarial or red-teaming cases. Candidate attacks include phone-assisted generation, memorized AI-generated text, manual retyping of an external AI answer, and GUI-agent imitation of human writing behavior. These cases are not part of the first final-text detector benchmark; they test the limits of process evidence itself. The trace taxonomy includes fully human typing, mostly human typing with some paste, human typing plus AI editing, AI brainstorming followed by human writing, AI-generated portions, and AI-generated full text.

\paragraph{Planned Outputs.}
The final paper will report writer usability, policy clarity, reader accuracy, confidence, and qualitative evidence about where certificate/replay helped or confused readers. \designnote{Finalize prompt pool, recruitment size, condition assignment, participant payment, and analysis plan.}

\section{Limitations and Ethical Considerations}
\label{sec:limitations}

\paragraph{Process evidence is not automatic proof.}
Humanly issues process evidence, not an automatic authorship verdict. A certificate can show typed characters, pasted characters, AI-use statistics, and replayable history, but a reader must still interpret that evidence in context.

\paragraph{Adversarial behavior.}
Humanly can block or log some actions, such as paste depending on task policy, but it cannot prevent every external-AI workflow. A determined user may transcribe external AI output, use voice input, mimic revisions, or route text through another device. The system should therefore be framed as evidence infrastructure, not as a complete anti-cheating mechanism.

\paragraph{Privacy and surveillance.}
Process recording is sensitive. It captures typing behavior, paste activity, focus changes, and AI interactions. Humanly's use should therefore be transparent to writers, limited to contexts where process evidence is appropriate, and paired with clear policies about who can inspect logs, certificates, and replays.

\paragraph{Study ethics.}
The planned studies will use non-sensitive prompts, disclose process recording, and avoid collecting personally identifying information beyond what is needed for participant payment. If any task involves mild deception, such as withholding details about paste behavior until debriefing, the protocol must justify the necessity and include debriefing. \designnote{Finalize ethics text once the actual study protocol is locked.}

\section{Conclusion}
\label{sec:conclusion}

\textsc{Humanly} reframes writing authenticity as a process-evidence problem. In the LLM era, the important question is not only whether a text looks AI-generated, but whether the writing process complied with the rules that governed the task. Humanly operationalizes this view by combining a tracked writing workspace, configurable AI access, task policy, and verifiable certificates. \designnote{Revise final paragraph after Section~\ref{sec:comparison} and Section~\ref{sec:study} contain completed evaluation results.}

% Bibliography entries for the entire Anthology, followed by custom entries
%\bibliography{anthology,custom}
% Custom bibliography entries only
\bibliography{custom}

% TODO: optional appendix, maximum two pages. Candidate content:
% - final prompt pool;
% - detector stress-test situation cases;
% - human-study survey items;
% - extra screenshots or certificate example.

\end{document}
